ForOpenCL: Transformations Exploiting Array Syntax in Fortran for Accelerator Programming

### 2011-06-27 papers/GPUScA-2011/programming-model.tex
spelling and ack

A question that one may pose is ``\emph{Why choose Fortran and not a more modern
  language like X for programming accelerator architectures?}''  The recent rise
in interest in concurrency and parallelism at the language level driven by
multicore CPUs and manycore accelerators has driven a number of new language
developments, both as novel languages and extensions on existing ones.  However,
for many scientific users with existing codes written in Fortran, new languages
and language extensions to use novel new architectures present a challenge: how
do they effectively use them while avoiding rewriting code and potentially
growing dependent on a transient technology that will vanish tomorrow?
In this paper we explore the constructs in Fortran that are particularly
relevant to GPU architecures.

In this section we present the Fortran subset employed in this paper.  This
subsetting language will allow scientific programmers to stay within the Fortran
language and yet have direct access to GPU hardware.  We start by examining how this
programming model relates to developments in other languages.

@@

A number of previous efforts have exploited data-parallel programming at the
language level to utilize novel architectures.
%%, particularly in previous decades during the reign of vector and massively parallel computers in high performance computing (HPC). 
The origin of the array syntax adopted by Fortran in the 1990
standard can be found in the APL language~\cite{iverson79apl}.  These additions
to Fortran allowed parallelism to be expressed with whole-array operations at
the expression level, instead of via parallelism within explicit DO-loops (e.g.,
IVTRAN for the Illiac IV).

>>>>

A question that one may pose is ``\emph{Why choose Fortran and not a more modern
  language like X for programming accelerator architectures?}''  The recent rise
in interest in concurrency and parallelism at the language level driven by
multicore CPUs and many-core accelerators has driven a number of new language
developments, both as novel languages and extensions on existing ones.  However,
for many scientific users with existing codes written in Fortran, new languages
and language extensions to use novel new architectures present a challenge: how
do they effectively use them while avoiding rewriting code and potentially
growing dependent on a transient technology that will vanish tomorrow?
In this paper we explore the constructs in Fortran that are particularly
relevant to GPU architectures.

In this section we present the Fortran subset employed in this paper.  This
sub-setting language will allow scientific programmers to stay within the Fortran
language and yet have direct access to GPU hardware.  We start by examining how this
programming model relates to developments in other languages.

@@

A number of previous efforts have exploited data-parallel programming at the
language level to utilize novel architectures.
%%, particularly in previous decades during the reign of vector and massively parallel computers in high performance computing (HPC). 
The origin of the array syntax adopted by Fortran in the 1990
standard can be found in the APL language~\cite{iverson79apl}.  These additions
to Fortran allowed parallelism to be expressed with whole-array operations at
the expression level, instead of via parallelism within explicit DO-loops, as
implemented in earlier variants of the language (e.g.,
IVTRAN for the Illiac IV).

### 2011-06-27 papers/GPUScA-2011/source-to-source.tex
spelling and ack

Device memory is separated hierarchically.  A thread instance has access to its
own thread memory (normally a set of registers), threads in a thread group to
OpenCL local memory, and all thread groups have access to OpenCL global memory.
When multiple members of a thread group access the same memory elements (for
example in the use of the {\tt region\_cpy} function in the calculation of face
variable quantites shown above), for performance reasons it is often best that
global memory accessed and shared by a thread group be copied into local memory.

>>>>

Device memory is separated hierarchically.  A thread instance has access to its
own thread memory (normally a set of registers), threads in a thread group to
OpenCL local memory, and all thread groups have access to OpenCL global memory.
When multiple members of a thread group access the same memory elements (for
example in the use of the {\tt region\_cpy} function in the calculation of face
variable quantities shown above), for performance reasons it is often best that
global memory accessed and shared by a thread group be copied into local memory.
